Scalable Extreme Deconvolution

### 2019-09-03 ml4ps_paper/paper.tex
Add R to EM equations.

The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
It requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.

@@

Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
For simplicity in the remainder of this work we will set $R_i$ to be the identity matrix and handle missing data via the covariance method.

@@

The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size N, using our current estimate of the parameters,
\begin{equation}
r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid \bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid \bm_k, T_{ik})}, \quad
\bb_{ij} = m_j + V_jT_{ij}^{-1}(\bx_i - \bm_j), \quad
B_{ij} = V_j - V_j T_{ij}^{-1}V_j.
\end{equation}
The $r_{ij}$ term is just the posterior probability of datapoint $\bx_i$ coming from component $j$.
The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
The expected sufficient statistics are then summed together over the minibatch,
\begin{equation}
\phi_j^1 = \sum_i r_{ij}, \quad
\phi_j^2 = \sum_i r_{ij} \bb_{ij}, \quad
\phi_j^3 = \sum_i r_{ij} [\bb_{ij}\bb_{ij}^T + B_{ij}].
\end{equation}
Stochastic estimates $\hat{\phi}_{tj} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
\begin{equation}
\hat{\phi}_{tj} = (1 - \lambda)\hat{\phi}_{(t-1)j} + \lambda \phi_{tj}.
\end{equation}
Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters.
\begin{equation}
\alpha_j = \frac{\hat{\phi}_j^1}{N}, \quad
\bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
\end{equation}
This procedure is repeated with new randomly ordered minibatches until convergence.
If we set $\lambda=1$ and replace each minibatch with the entire dataset then we recover the original fitting method.

\todo{Discuss numerical stability of the $V_j$ update.}

@@

We implemented both of our methods in PyTorch.
To evaluate our methods, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
In total there were 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used the same number of components $K=128$ for each model, following \citet{andersonImprovingGaiaParallax2018}.
In practice we would want to select a value of $K$ by cross-validation.

Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
Training was done for a maximum of 20 epochs.
Early stopping was done based on a held-out validation set comprising 10\% of the rows.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection performed based on this value.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.

@@

\begin{table}
  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Method     & Validation     & Test & Training Time (minutes) \\
    \midrule
    Existing EM & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
    Minibatch EM & $-5.183 (\pm 0.0005) \times 10^6$    & - & $48.7 \pm 0.39$      \\
    SGD     & $-5.193 (\pm 0.0029) \times 10^6$        & - & $36.0 \pm 0.34$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

>>>>

The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
It must make a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.

@@

Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.

@@

The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size N, using our current estimate of the parameters,
\begin{equation}
r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
\end{equation}
The $r_{ij}$ term is just the posterior probability of datapoint $\bx_i$ coming from component $j$.
The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
The expected sufficient statistics are then summed together over the minibatch,
\begin{equation}
\phi_j^1 = \sum_i r_{ij}, \quad
\phi_j^2 = \sum_i r_{ij} \bb_{ij}, \quad
\phi_j^3 = \sum_i r_{ij} [\bb_{ij}\bb_{ij}^T + B_{ij}].
\end{equation}
Stochastic estimates $\hat{\phi}_{tj} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
\begin{equation}
\hat{\phi}_{tj} = (1 - \lambda)\hat{\phi}_{(t-1)j} + \lambda \phi_{tj}.
\end{equation}
Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters.
\begin{equation}
\alpha_j = \frac{\hat{\phi}_j^1}{N}, \quad
\bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
\end{equation}
This procedure is repeated with new randomly ordered minibatches until convergence.
If we set $\lambda=1$, replace each minibatch with the entire dataset and factorise the update for $V_j$ then we recover the original fitting method.
The update for $V_j$ could result in catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
In Appendix~\ref{apx:variance-rewrite} we show how this update can be rewritten in a symbolically equivalent but more numerically stable form.

@@

We implemented both of our methods in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
In total there were 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
We set the projection $R_i$ to the identity matrix for every sample.
This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used the same number of components $K=256$ for each model.
In practice we would want to select a value of $K$ by cross-validation.

Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
Training was done for 20 epochs.
We used a validation set comprising $10\%$ of the rows when developing our experiments.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.

@@

\begin{table}
  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Method     & Validation     & Test & Training Time (minutes) \\
    \midrule
    Existing EM (\citet{bovyExtremeDeconvolutionInferring2011}) & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
    Minibatch EM & $-5.183 (\pm 0.0005) \times 10^6$    & - & $48.7 \pm 0.39$      \\
    SGD     & $-5.193 (\pm 0.0029) \times 10^6$        & - & $36.0 \pm 0.34$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\appendix

\section{Stable Variance Update}
\label{apx:variance-rewrite}

\section{Experiment Details}
